Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-05-14 Manuscript/manuscript.Rmd
manuscript edits

To establish the data for modeling, the genome was binned into equally sized intervals defined by the data resolution for the corresponding contact matrix (Figure 1). As an example, for domain data obtained from a 10 kb resolution contact matrix, the genome was binned into 10 kb intervals. For the boundary data, in order to accommodate for the uncertainty in exact boundary location, the boundary points were flanked on both sides by the size of the resolution of the corresponding contact matrix. That is, for 10 kb resolution data, each uniquely identified TAD boundary was flanked on either side by 10 kilobases for a total width around each boundary point of 20 kb. A genomic bin was then labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing a boundary (Y=0). The process was repeated for each subsequent resolution.

@@

Using the sets of genomic annotations, we created three types of predictors that were used as features for subsequent models to analyze their respective spatial relationship with TAD boundaries (Figure 3):

@@

3. **Log2 Distance**: For each genomic bin, the $log_{2}$ of the distance in bases (refered to hereafter simply as $\textit{distance}$), from the center of the bin to the center of the nearest region defined by a specific genomic annotation was calculated. The log transformation was used as a normalization technique in order to account for the skewness present in the distance densities (Supplementary Figures 1).

@@

4. **SMOTE** (Synthetic Minority Over-Sampling Technique): This method incorporates both random under- and over-sampling. Under-sampling is performed without replacement from the majority class, while over-sampling is performed by creating new synthetic observations using the minority class [@chawla2002smote]. The `DMwR` R package was used to perform SMOTE (version 0.4.1).

<!--
5. **ROSE** (Random Over-Sampling Examples): Similar to SMOTE, ROSE combines techniques of over-sampling and under-sampling. However, contrary to SMOTE, all of the data generated by ROSE is synthetic. The data is created by sampling from one of the two classes, then generating a new example in its neighborhood using a smoothed bootstrap approach. The shape of the neighborhood is determined by the kernel density estimate of the two classes [@lunardon2013r]. Another important distinction between ROSE and SMOTE is that, for ROSE, the creation of perfectly balanced classes is unlikely due to the randomness in sampling between the two classes. The `ROSE` R package was used to perform ROSE (version 0.0.3).
-->

@@

A regularization technique using a combination of the $l_{1}$ and $l_{2}$ penalties, known as the elastic-net, was used to reduce the feature space. The elastic-net was chosen in order to suitably reduce the feature space while also maintaining the inclusion of multiple correlated predictors. For each elastic-net regularization, 10-fold cross-validation was used to tune the parameters that controlled the penalty terms. The parameters that maximized the mean AUPRC across each cross-fold was chosen. All elastic-net algorithms were implemented in R using the `caret` package version 6.0 [@kuhn2012caret].

<!--
Recursive feature elimination (RFE) was then used to determine the optimal number of features for models with the best performing predictor type. For each set of predictor type, random forest models were performed recursively on subsets of the feature space. The specific subsets of features were chosen as powers of 2 for computational efficiency. Model performance for each subset was evaluated using AUPRC. The optimal set of featues to include in downstream analyses was determined to be the subset in which the AUPRC leveled off. Again, to confirm that results were consistent, the reduction process was repeated for each combination of re-sampling technique, resolution, and cell line. 
-->

@@

Once the appropriate predictor type and set of features were determined, a final set of random forest classification models were built to compare model performances between re-sampling techniques and predictor types. <!--To reduce bias due to random dataset generation, 10-fold cross validation was used to tune the number of features to consider at each node and the number of ensemble trees to aggregate.--> The default number of features to consider at each node of the random forest algorithm was set as $\sqrt{p}$, where $p$ refers to the number of predictors included in the model. Likewise, the number of ensemble trees to aggregate was set at 500. All random forest algorithms were implemented in R using the `caret` package version 6.0 [@kuhn2012caret]. Once the model was implemented, it was validated on the test set.

@@

An ensemble framework was then established for the purposes of systematically creating a variety of models that each included different sets of parameters described above (Figure 2). Each system in the framework is assumed to be composed of a model constructed from a particular cell line and data resolution, with a particular predictor type for the feature space. The data sets used for modeling were composed of the full set of genomic bins that either overlapped with a flanked TAD boundary or did not. The data was then split into a 7:3 ratio of training to testing sets. For the task of performance evaluation and prediction, the test set remained constant for all ensemble systems to account for an unbiased comparison. Furthermore, this process was repeated, and performances were aggregated across each autosomal chromosome (omitting chromosome 9) using means and standard deviations.

@@

@@

<!--
Additionally, when we performed and evaluated the variable importances of random forest models built on one genomic annotation at a time (including the 3 predictor types each), we found that the distance-type consistently out-ranked the OC and OP type predictors in terms of predictive importance, for a majority of the cell line specific annotations, across resolution and cell line (Supplementary Tables 4 & 5). Therefore, it appears that distance-type predictors provided the most informative spatial representation of epigenomic annotations associated with TAD boundaries.
-->

>>>>

To establish the data for modeling, the genome was binned into equally sized intervals defined by the data resolution for the corresponding contact matrix <!--(Figure 1)-->. As an example, for domain data obtained from a 10 kb resolution contact matrix, the genome was binned into 10 kb intervals. For the boundary data, in order to accommodate for the uncertainty in exact boundary location, the boundary points were flanked on both sides by the size of the resolution of the corresponding contact matrix. That is, for 10 kb resolution data, each uniquely identified TAD boundary was flanked on either side by 10 kilobases for a total width around each boundary point of 20 kb. A genomic bin was then labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing a boundary (Y=0). The process was repeated for each subsequent resolution.

@@

Using the sets of genomic annotations, we created three types of predictors that were used as features for subsequent models to analyze their respective spatial relationship with TAD boundaries <!--(Figure 3)-->:

@@

3. **Log2 Distance**: For each genomic bin, the $log_{2}$ of the distance in bases (refered to hereafter simply as $\textit{distance}$), from the center of the bin to the center of the nearest region defined by a specific genomic annotation was calculated. The log transformation was used as a normalization technique in order to account for the skewness present in the distance densities <!--(Supplementary Figures 1)-->.

@@

4. **SMOTE** (Synthetic Minority Over-Sampling Technique): This method incorporates both random under- and over-sampling. Under-sampling is performed without replacement from the majority class, while over-sampling is performed by creating new synthetic observations using the minority class [@chawla2002smote]. The *DMwR* R package was used to perform SMOTE (version 0.4.1).

@@

A regularization technique using a combination of the *l_{1}* and *l_{2}* penalties, known as the elastic-net, was used to reduce the feature space. The elastic-net was chosen in order to suitably reduce the feature space while also maintaining the inclusion of multiple correlated predictors. For each elastic-net regularization, 10-fold cross-validation was used to tune the parameters that controlled the penalty terms. The parameters that maximized the mean AUPRC across each cross-fold was chosen. All elastic-net algorithms were implemented in R using the *caret* package version 6.0 [@kuhn2012caret].

@@

Once the appropriate predictor type and set of features were determined, a final set of random forest classification models were built to compare model performances between re-sampling techniques and predictor types. <!--To reduce bias due to random dataset generation, 10-fold cross validation was used to tune the number of features to consider at each node and the number of ensemble trees to aggregate.--> The default number of features to consider at each node of the random forest algorithm was set as $\sqrt{p}$, where *p* refers to the number of predictors included in the model. Likewise, the number of ensemble trees to aggregate was set at 500. All random forest algorithms were implemented in R using the *caret* package version 6.0 [@kuhn2012caret]. Once the model was implemented, it was validated on the test set.

@@

An ensemble framework was then established for the purposes of systematically creating a variety of models that each included different sets of parameters described above <!--(Figure 2)-->. Each system in the framework is assumed to be composed of a model constructed from a particular cell line and data resolution, with a particular predictor type for the feature space. The data sets used for modeling were composed of the full set of genomic bins that either overlapped with a flanked TAD boundary or did not. The data was then split into a 7:3 ratio of training to testing sets. For the task of performance evaluation and prediction, the test set remained constant for all ensemble systems to account for an unbiased comparison. Furthermore, this process was repeated, and performances were aggregated across each autosomal chromosome (omitting chromosome 9) using means and standard deviations.

@@

## THE PROPOSED MODEL

We illustrate a unified approach toward preparing TAD boundary data for predictive modelling purposes from different cell lines and data resolutions (Figure 1). To enable us to build a model that suitably discriminates between regions on the linear genome that are near TAD boundaries and regions that are not, we created a cell line- and resolution-specific response variable that indicated whether a genomic bin of unit width overlapped with a flanked TAD boundary (Y=1), or not (Y=0). The unit width of each bin on the linear genome corresponded to the resolution of the contact matrix that domains were called on. Moreover, to account for uncertainty in boundary location, TAD boundaries were flanked on either side by the same unit length as the genomic bins.

To investigate the relationship between functional genomic annotations and the development of TAD boundaries, we devised 3 different predictor-types that composed the feature space of downstream random forest classification models. These predictor-type specific annotations represented the data matrix, **X** = {$X_{1}$, $X_{2}$, ..., $X_{p}$}, where *p* the set of annotations. Overlap counts was the least granular predictor-type considered, simply defined by the number of annotation-specific regions located within each genomic bin (Figure 2A). Overlap percents was a more refined predictor-type, defined as the total percent overlap between each annotation-specific region and genomic bins (Figure 2B). Lastly, distances, the most granular predictor-type, described the spatial relationship between the center of each genomic bin to the center of the nearest annotation-specific region (Figure 2C).  When measured in basepairs, distances were found to be highly skewed across all resolutions, for both the GM12878 and K562 cell lines (Supplementary Figures 1-2 A-D). As a normalization procedure, we performed a $log_{2}$ transformation on the distances (Supplementary Figures 1-2 E-H).

@@
